Social Participation Ontology: community documentation, enhancements and use examples

### 2015-01-09 paper.tex
adicionado uso dia dereferenciamento

\date{\today}% It is always \today, today,
             %  but any date may be explicitly specified
\newcommand{\ops}{\textsc{ops}}
\newcommand{\opsi}{O\textsc{ps}}
\newcommand{\vcps}{\textsc{vcps}}
\newcommand{\owl}{\textsc{owl}}
\newcommand{\sparql}{\textsc{s}par\textsc{ql}}
\newcommand{\bfo}{\textsc{bfo}}
\newcommand{\dbpedia}{\textsc{db}pedia}
\newcommand{\foaf}{\textsc{foaf}}
\newcommand{\ict}{\textsc{ict}}
\newcommand{\html}{\textsc{html}}
\newcommand{\node}{\textsc{n}ode.js}
\newcommand{\facebook}{\textsc{f}acebook}
\newcommand{\twitter}{\textsc{t}witter}
\newcommand{\wwwc}{\textsc{w3c}}
\newcommand{\skos}{\textsc{skos}}
\newcommand{\etherpad}{\textsc{e}therpad}
\newcommand{\ogp}{\textsc{ogp}}
\newcommand{\iri}{\textsc{iri}}
\newcommand{\uri}{\textsc{uri}}
\newcommand{\urll}{\textsc{url}}
\newcommand{\ngo}{\textsc{ngo}}
\newcommand{\http}{\textsc{http}}
%\newcommand{\t}[1]{\texttt{#1}}
\begin{abstract}
    Participatory democracy advances in virtually all governments. South America presents a prominent context with mixed culture and social predisposition. At 2012, civil, academic and governmental parties started elaborating the ``Common Vocabulary of Social Participation'' (\vcps\ from the Brazilian name \emph{Vocabul\'ario Comum de Participa\c{c}\~ao Social}), as a public and online process. By May, 2013, first reference documents were publicized, together with a preliminary \owl\ code,
logos, and a diagram of a general ``public consultation''.
The Corais Platform kept online records of the process, like discussions and preparation of texts. 
This article exposes this material and proposes elementary unfolding, highlighting the ``Social Participation Ontology'' (\ops\ from the Brazilian name \emph{Ontologia de Participa\c{c}\~ao Social}). To exhibit this new ontology, this steps were considered: completion of \vcps\ \owl\ code for preliminary \ops, with necessary translations and corrections; correction of ontological contradictions and enhancements of class names and labels in Portuguese, Spanish and English; a toy expansion of the ontology by further specifying classes; use cases of \ops\ by researchers and public managers; and use examples regarding knowledge organization and a \sparql\ endpoint. Ongoing work involves incorporation by the official Brazilian social participation federal portal, incorporation by civil participative organizations, linkage to other participative ontologies. \ops\ is being used as an upper ontology, and all classes linked to \foaf\ and \bfo\ as higher upper ontologies.
\end{abstract}

@@

Next section is dedicated to an inspection of the documents delivered by the \vcps: reference textual documents, images, OWL code, blog posts, discussions and \etherpad s. Section~\ref{exp} presents \ops: modifications made from \vcps\ to \ops: class and property names and labels, class restrictions and property axioms. Section~\ref{ospUtil} is dedicated to \ops\ usage: derreferatiation, \sparql\ queries, a toy \ops\ expansion, discursive example of usage, and use cases from government, civil society and academic parties. Concluding remarks are stated with future work, in Section~\ref{conc}.

@@

\section{OSP utility}\label{ospUtil}
OSP is meant to be useful. First, as a sistematization of what is social participation to Latin America groups, as conceived by CVSP. Second, as a mean to ease linked data, and enable integration of various instances for social participation. An indicative of this pertinence is OPA, an ontology that already uses OSP.

\subsection{Usage of OSP via a SparQL endpoint}
Accessing multiple databases is one of the OWL ontology basic purposes. The standard way to acess these data via ontologies is by using a SparQL endpoint. This endpoint delivers data from a triplestore or, with more experimental technology, relational database systems, such as a MySQL server (e.g. via OnTop/Quest~\cite{onTop}). Either way, the query is the same, i.e. the user or machine reaching the endpoint uses the same protocol in order to retrieve information through semantic criteria. Figure~\ref{endpoint} is a schematic representation of OBDA (Ontology Based Database Access), which is a common name for this multiple database access through ontologies.

>>>>

\date{\today}% It is always \today, today,
             %  but any date may be explicitly specified
\newcommand{\ops}{\textsc{ops}}
\newcommand{\opsi}{O\textsc{ps}}
\newcommand{\vcps}{\textsc{vcps}}
\newcommand{\owl}{\textsc{owl}}
\newcommand{\sparql}{\textsc{s}par\textsc{ql}}
\newcommand{\bfo}{\textsc{bfo}}
\newcommand{\dbpedia}{\textsc{db}pedia}
\newcommand{\foaf}{\textsc{foaf}}
\newcommand{\ict}{\textsc{ict}}
\newcommand{\html}{\textsc{html}}
\newcommand{\node}{\textsc{n}ode.js}
\newcommand{\facebook}{\textsc{f}acebook}
\newcommand{\twitter}{\textsc{t}witter}
\newcommand{\wwwc}{\textsc{w3c}}
\newcommand{\skos}{\textsc{skos}}
\newcommand{\etherpad}{\textsc{e}therpad}
\newcommand{\ogp}{\textsc{ogp}}
\newcommand{\iri}{\textsc{iri}}
\newcommand{\uri}{\textsc{uri}}
\newcommand{\urll}{\textsc{url}}
\newcommand{\ngo}{\textsc{ngo}}
\newcommand{\http}{\textsc{http}}
\newcommand{\opa}{\textsc{op}a}
\newcommand{\ocd}{\textsc{ocd}}
\newcommand{\ontologiaa}{\textsc{o}ntologiaa}
\newcommand{\obs}{\textsc{obs}}
\newcommand{\pubby}{\textsc{p}ubby}
\newcommand{\rdf}{\textsc{rdf}}
%\newcommand{\t}[1]{\texttt{#1}}
\begin{abstract}
    Participatory democracy advances in virtually all governments. South America presents a prominent context with mixed culture and social predisposition. At 2012, civil, academic and governmental parties started elaborating the ``Common Vocabulary of Social Participation'' (\vcps\ from the Brazilian name \emph{Vocabul\'ario Comum de Participa\c{c}\~ao Social}), as a public and online process. By May, 2013, first reference documents were publicized, together with a preliminary \owl\ code,
logos, and a diagram of a general ``public consultation''.
The Corais Platform kept online records of the process, like discussions and preparation of texts. 
This article exposes this material and proposes elementary unfolding, highlighting the ``Social Participation Ontology'' (\ops\ from the Brazilian name \emph{Ontologia de Participa\c{c}\~ao Social}). To exhibit this new ontology, this steps were considered: completion of \vcps\ \owl\ code for preliminary \ops, with necessary translations and corrections; correction of ontological contradictions and enhancements of class names and labels in Portuguese, Spanish and English; a toy expansion of the ontology by further specifying classes; use cases of \ops\ by researchers and public managers; and use examples regarding knowledge organization and a \sparql\ endpoint. Ongoing work involves incorporation by the official Brazilian social participation federal portal, incorporation by civil participative organizations, linkage to other participative ontologies. \ops\ is being used as an upper ontology, and all classes linked to \foaf\ and \bfo\ as higher upper ontologies.
\end{abstract}

@@

Next section is dedicated to an inspection of the documents delivered by the \vcps: reference textual documents, images, OWL code, blog posts, discussions and \etherpad s. Section~\ref{exp} presents \ops: modifications made from \vcps\ to \ops: class and property names and labels, class restrictions and property axioms. Section~\ref{ospUtil} is dedicated to \ops\ usage: dereferencing, \sparql\ queries, a toy \ops\ expansion, discursive example of usage, and use cases from government, civil society and academic parties. Concluding remarks are stated with future work, in Section~\ref{conc}.

@@

\section{\ops\ utility}\label{ospUtil}
\opsi\ is meant to be useful. First, as a systematization of what is social participation to Latin America groups, as conceived by \vcps. Second, as a mean to ease linked data, and enable integration of various instances for social participation. An indicative of this pertinence is \opa~\cite{opa}, \ocd~\cite{ocd}, \ontologiaa~\cite{ontologiaa}, and \obs~\cite{obs}, ontologies that already uses \ops\ as upper ontology.

This section explores different \ops\ uses: dereferencing, \sparql\ queries, expansion, discursive fictional cases, and real use cases.

\subsection{Usage of \ops\ through dereferencing}

All \ops\ classes and properties \uri s are accessible via \http. A \pubby\ instance delivers information like name, labels and relations to other classes and properties. As an example, the \uri\ \url{http://purl.org/socialparticipation/ops/SocialActor} returns information about this class and all subclasses it is related to, as shown in Figure~\ref{fig:deref}.

\begin{figure}[!h]
    \centering
    %\includegraphics[width=\columnwidth]{figs/pubbyDer}
    \includegraphics[width=\columnwidth]{figs/foo}
    \caption{Dereferencing an \ops\ class: the \uri\ is also an \urll, which, reached by \http, delivers information for the user as shown. Also, if the client is not a browser, but a crawler or a linked data application, \pubby\ delivers plain \rdf, not user-friendly \html.}
    \label{fig:owlCC}
\end{figure}

\subsection{Usage of \ops\ through a \sparql\ endpoint}
Accessing multiple databases is one of the OWL ontology basic purposes. The standard way to acess these data via ontologies is by using a SparQL endpoint. This endpoint delivers data from a triplestore or, with more experimental technology, relational database systems, such as a MySQL server (e.g. via OnTop/Quest~\cite{onTop}). Either way, the query is the same, i.e. the user or machine reaching the endpoint uses the same protocol in order to retrieve information through semantic criteria. Figure~\ref{endpoint} is a schematic representation of OBDA (Ontology Based Database Access), which is a common name for this multiple database access through ontologies.
